 

### 2021-06-25 vertical-thermal-review.Rmd
start at major revamp of leaf T section
- add par on leaf energy budget
- add par on how biophysical principle shape Tleaf across vert gradient
-propose cutting a lot of content

@NidhiVinod , I have to take a break now, but hope to come back to this soon.

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, a optimal for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ (poikilothermy) [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020]. 
Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}$-$T_{air}$ is a function of energy input (through radiation or heat) versus heat lost to the environment (such as transpiration) [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].

@@

*(In general, we'll want to adjust this section to parallel the figure and tie in with the section above.)*

@@

@@

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].

>>>>

Leaf temperature ($T_{leaf}$) is strongly tied to $T_{air}$, and is further influenced by the other biophysical drivers reviewed above and by leaf traits (see next section), with leaves adjusted to their environment so as to maintain, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. 
Leaves can be cooler than air at high $T_{air}$ and warmer than air at low $T_{air}$ (limited homeothermy), and may sometimes be drastically warmer than $T_{air}$ (megathermy).
Often, however, $T_{leaf}$ tracks $T_{air}$ [poikilothermy, @cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

**Fundamentally, $T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles, where $T_{leaf}-T_{air}$ is a function of energy input through radiation versus heat lost to the environment [Fig. `r fig_leaf_T`, @campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016; @michaletzPlantThermoregulationEnergetics2015].**
The dominant component of a leaf's energy balance is net radiation ($R_n$, including shortwave and longwave), and high radiation loads can elevate $T_{leaf}$ dramatically above $T_{air}$ (Fig. `r fig_leaf_T`a).
The dominant avenue of heat loss to the environment is latent heat flux ($\lambda E$), wherein the loss of energy required convert water from liquid to vapour state cools the leaf. 
Evaporation ($E$) is controlled primarily by stomatal and boundary layer conductances, $g_s$ and $g_a$, and by the water vapour concentration difference between the leaf surface and the bulk air.
Therefore, $T_{leaf}-T_{air}$ decreases with $g_a$, which in turn increases with wind speed (Fig. `r fig_leaf_T`b) and decreases with leaf size (Fig. `r fig_leaf_T`d);
decreases with $g_s$ (Fig. `r fig_leaf_T`e);
and decreases with with vapor pressure deficit, or increases with RH (Fig. `r fig_leaf_T`c).

@@

**These basic biophysical principles shape the propensity for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in forests.**
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater propensity for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher windspeeds and lower RH (Fig. `r fig_NEON_vertical`), along with adaptive leaf traits (see next section), give them lower sensitivity to intercepted radiation (Fig. `r fig_leaf_T`a) and higher rates of heat loss (Fig. `r fig_leaf_T`b-e).
*(I want to add a few sentences on how this varies with environmental conditions-- drought, high humidity....)*

~~~ (*I think we should cut everything between here and "In-situ observations", after harvesting the most important content for the section above.*)

@@

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. In contrast, when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissipation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].
